The Design of an Aliasing-Tolerant Parallel Programming Model

### 2012-11-08 doc/types-pldi2013/alt_example.tex
bunch o' changes to section 2

The simulation takes as input an arbitrary graph of circuit elements
that have been dynamically allocated into two logical regions: {\tt all\_nodes} 
and {\tt all\_wires}.  The types of nodes and wires are the tuples
shown in Listing~\ref{lst:tuples}.  The first two fields in the
{\tt CircuitWire} tuple are pointers that specify the two end points of
a wire.  Pointers are denoted by the @ symbol
followed by the set of region(s) into which a pointer can point.
The {\tt CircuitWire} type is parametrized by the local region
names {\tt rn} and {\tt rg} which contain {\tt CircuitNode} elements.
Our type system enforces the correctness of pointer types.

@@

The most important decision in any Legion program is the choice of
how to partition program data.  The choice of a partitioning algorithm
can depend on a multitude of factors including the type of data structure
and input data.  Therefore
Legion requires applications to select their own partitioning algorithms.
In this example we invoke a third-party dynamic graph partitioning library
called METIS\cite{Metis98} as shown in Listing~\ref{lst:metis}.

@@

METIS takes as arguments lists of all the nodes and all the wires in
the graph.  It then breaks the graph into pieces by annotating each 
{\tt CircuitNode} with a piece ID (the last field in the tuple).
The function {\tt extern\_metis} is parametrized on the regions it will
access; in this case {\tt rn} the region containing nodes, {\tt rw} the region
contining wires, and {\tt rl} the region containing the lists of pointers.  The function
must also specify the {\em privileges} (e.g read) with which it will access
these regions.  {\tt extern\_metis} will read all three regions and write
the {\tt rn} region since it will be writing to the last field of each
{\tt CircuitNode}.  Our type system enforces the Legion 
requirement that each function can only be invoked with a subset of the 
privileges contained in its calling context.

@@

Figure~\ref{sfig:part_fig:pieces} shows one possible output from
METIS with different colors identifing different piece IDs.  To capture
the result of a partitioning algorithm, Legion uses {\em colorings}.
Colorings are a mapping from pointers to {\em colors} (e.g. integers).  In
the circuit example, the piece IDs are the designated colors. 
Listing~\ref{lst:coloring} shows the function {\tt create\_piece\_coloring}
that constructs a coloring from a list of nodes which have been 
annotated with piece IDs.

@@

Colorings are the input for partition operations in Legion.  The partition
statement shown in Listing~\ref{lst:partition} partitions the region
{\tt all\_nodes} into logical subregions with a separate output region
for each color in the coloring.  Due to space constraints we, only show
code for {\tt rn0} and {\tt rn1} pieces for the remainder of the example.  We also
do not show the coloring or partitioning of the {\tt all\_wires} region as
it can be computed in a straightforward manner from the partitioning of
nodes into pieces.

@@

Since colorings can only color each pointer at most with a single color
(dynamically checked), the partition operation results in 
{\em disjoint} subregions.  Furthermore, the partition operation
introduces constraints into the static checking enviornment describing
both the disjointness of the subregions (i.e. {\tt rn0 $*$ rn1})
and the inclusion of each subregion in the original region
(i.e. {\tt rn0 $\leq$ all\_nodes}).

In addition, to partitioning the graph to describe {\em owned} nodes
for each piece, the circuit example also
needs to be able to name the {\em ghost nodes} for each piece.  Ghost node
regions are used for accessing the nodes of wires that span
two different pieces. (This is also the reason {\tt CircuitWire} is parametrized on
two regions).  Figure~\ref{sfig:part_fig:ghost_zero} and 
Figure~\ref{sfig:part_fig:ghost_one} show colorings for the ghost regions
corresponding to {\tt rn0} and {\tt rn1} respectively.  Note that two 
nodes are colored in both colorings.  Since nodes can appear in multiple
subregions, we use a {\em multicoloring} (implemented in core Legion
by multiple single colorings) to describe {\em aliased} partitions.  Aliased
partitions still introduce subregion constraints into the static checking
environment, but not disjointness constraints.  Figure~\ref{sfig:part_fig:tree}
shows the resulting logical region tree for node regions.

The first-class nature of logical regions in Legion allows them to
be written into the heap.  To avoid losing important static properties of
logical regions in this scenario, we use {\em region
relationships}.  Region relationships are a bounded existential type
that allows programmers to {\em pack} a group of regions together and remember
properties about them such as disjointness and subregion relationships.
Listing~\ref{lst:rr} shows an example region relationship for a {\tt CircuitPiece} 
describing its region of wires {\tt rpw}, region of 
nodes {\tt rpn}, region of ghost nodes {\tt rg}, and important constraints.

@@

The Legion type system verifies properties of region relationships when packing;
when region relationships are unpacked the same properties are
reintroduced into the static checking environment.
A key feature underlying the soundness of the Legion
type system is that privileges cannot be packed in a region
relationship---privileges belong to functions. When a function unpacks
a region relationship it must already hold privileges for the unpacked regions.
Listing~\ref{lst:loop} shows the main time step loop run by the 
recursive function {\em execute\_time\_steps} (line 11)
as well the leaf task functions (lines 2-7).  
For each time step, the loop unpacks 
two previously packed circuit pieces, giving new names to the subregions introduced
by each region relationship.  The {\em execute\_time\_steps} function
will have read/write privileges for the newly named regions, such as {\tt rn0},
because it has read/write privileges for {\tt rn} and the {\tt CircuitPiece} 
region relationship declares {\tt rn0 $\leq$ rn}.  Note this does require
a dynamic check that {\tt rn0} is a subregion of {\tt rn}, but the cost of the 
check will be ammortized over all the uses of {\tt rn0}.

>>>>

The simulation takes as input an arbitrary graph of circuit elements
that have been dynamically allocated into two logical regions: {\tt all\_nodes} 
and {\tt all\_wires}.  The types of nodes and wires are the tuples
shown in Listing~\ref{lst:tuples}.  The first two fields in the
{\tt CircuitWire} tuple are pointers that specify the two end points of
a wire.  Pointers are denoted by the @ symbol
followed by the region(s) into which the pointer may point.
The {\tt CircuitWire} type is parametrized by the local region
names {\tt rn} and {\tt rg} which contain {\tt CircuitNode} elements.
Our type system enforces the correctness of pointer types.

@@

The most important decision in any Legion program is the choice of
how to partition program data.  The ideal partitioning algorithm
depends on a multitude of factors including the type of data structure,
the input data, and the desired number of partitions (which often depends on the
size of the machine running the application).  Therefore
Legion permits applications to select their own partitioning algorithms.
In this example we invoke a third-party dynamic graph partitioning library
called METIS\cite{Metis98} as shown in Listing~\ref{lst:metis}.

@@

METIS takes as arguments lists of all the nodes and all the wires in
the graph.  It then breaks the graph into pieces by annotating each 
{\tt CircuitNode} with a piece ID (the last field in the tuple).
The function {\tt extern\_metis} is parameterized on the regions it will
access; in this case {\tt rn} is the region containing nodes, {\tt rw} is the region
contining wires, and {\tt rl} is the region containing the lists of pointers.  The function
must also specify the access {\em privileges} (e.g read, write) it requires for each of
these regions.  All three regions will be read by {\tt extern\_metis}, but only the {\tt rn}
region will be written to (since it will be modifying the last field of each {\tt CircuitNode}).
A function may only be invoked if the caller possesses all the privileges needed by the function
being called.  This is verified statically by the Legion type system.

@@

To efficiently support the access patterns of the circuit simulation,
multiple partitionings of the {\tt all\_nodes} region are performed.
The desired {\em region tree} is shown in Figure~\ref{sfig:part_fig:tree}.
First, we need subregions that describe the set of nodes ``owned'' by
each piece.  These will be called {\tt rn0}, {\tt rn1}, ... and since each
node has a single owner, they form the {\em disjoint} partition (indicated with a *)
on the left side of the tree.
Figure~\ref{sfig:part_fig:pieces} shows one possible output from
METIS with different colors identifing different piece IDs.
Second, each piece of the circuit will need to 
access the nodes that border its own.  These are commonly referred as {\em ghost nodes},
and the set of ghost nodes for the first two circuit pieces (named {\tt rg0} and {\tt rg1})
are shown in Figures~\ref{sfig:part_fig:ghost_zero}
and~\ref{sfig:part_fig:ghost_one}.  Because a node may neighbor more than one other circuit
piece, this second partition is an {\em aliased} partition.
Without the ability to have multiple partitionings of the same data, one would have two
possible alternatives:
\begin{enumerate}
\item Make a single partition with $2^N$ subregions, one for each possible case of sharing, or
\item Have all pieces use the {\tt all\_nodes} for their ghost cell region, greatly increasing the 
amount of data movement required
\end{enumerate}

To describe
the result of a partitioning algorithm, Legion uses {\em colorings}.
A coloring is a mapping from pointers to {\em colors} (represented in Core Legion as integers).  In
the circuit example, the piece IDs are the designated colors. 
Listing~\ref{lst:coloring} shows the function {\tt create\_piece\_coloring}
that constructs a coloring from a list of nodes which have been 
annotated with piece IDs.  The coloring of wires is similar, with each wire
using the color of its first endpoint.  For the ``ghost nodes'', a {\em multicoloring}
is used, which allows each pointer to be assigned multiple colorings.  (Multicolorings are
implemented in Core Legion with separate colorings for each color.)

@@

Colorings are the input for partition operations in Legion.  The partition
statement shown in Listing~\ref{lst:partition} performs the ``owned nodes''
partitioning.  The partitioning of wires is identical, while the partitioning
of the ghost nodes is performed in Core Legion as a separate partitioning operation
for each color.

@@

Since colorings can only color each pointer at most with a single color
(dynamically checked), the partition operation results in 
{\em disjoint} subregions, each of which contains a subset of the original region.
The partition operation introduces constraints into the type system describing
both the disjointness of the subregions (e.g. {\tt rn0 $*$ rn1})
and the inclusion of each subregion in the original region
(e.g. {\tt rn0 $\leq$ all\_nodes}).

%% In addition, to partitioning the graph to describe {\em owned} nodes
%% for each piece, the circuit example also
%% needs to be able to name the {\em ghost nodes} for each piece.  Ghost node
%% regions are used for accessing the nodes of wires that span
%% two different pieces. (This is also the reason {\tt CircuitWire} is parametrized on
%% two regions).  Figure~\ref{sfig:part_fig:ghost_zero} and 
%% Figure~\ref{sfig:part_fig:ghost_one} show colorings for the ghost regions
%% corresponding to {\tt rn0} and {\tt rn1} respectively.  Note that two 
%% nodes are colored in both colorings.  Since nodes can appear in multiple
%% subregions, we use a {\em multicoloring} (implemented in core Legion
%% by multiple single colorings) to describe {\em aliased} partitions.  Aliased
%% partitions still introduce subregion constraints into the static checking
%% environment, but not disjointness constraints.  Figure~\ref{sfig:part_fig:tree}
%% shows the resulting logical region tree for node regions.

The first-class nature of logical regions in Legion allows them to
be written into the heap.  To avoid losing information about these regions and pointers
into those regions, Legion provides {\em region
relationships}.  Region relationships are a bounded existential type
that allows programmers to {\em pack} a group of regions and pointers together and capture
properties about them such as disjointness and subregion relationships.
Listing~\ref{lst:rr} shows an example region relationship for a {\tt CircuitPiece} 
describing its region of wires {\tt rpw}, region of 
nodes {\tt rpn}, region of ghost nodes {\tt rg}, and important constraints.  Although the
``names'' of {\tt rpw}, {\tt rpn}, and {\tt rg} are lost when packed into the region relationship,
the knowledge that they are subregions of the {\tt rw} and {\tt rn} regions is not.

@@

The Legion type system verifies properties of region relationships when packing.  When
a region relationship is unpacked later, fresh names are given to the regions that were
captured by the region relationship, and the properties that are known to hold for these
regions are re-introduced to the type environment.
In constrast, region access privileges cannot be packed in a region
relationship---privileges belong to functions. When a function unpacks
a region relationship it must already hold privileges for the unpacked regions it wishes to access.

Listing~\ref{lst:loop} shows the main time step loop run by the 
recursive function {\em execute\_time\_steps} (line 11)
as well the leaf task functions (lines 2-7).  
For each time step, the loop unpacks 
two previously packed circuit pieces, giving new names to the subregions introduced
by each region relationship.  The {\em execute\_time\_steps} function
will have read/write privileges for the newly named regions, such as {\tt rn0},
because it has read/write privileges for {\tt rn} and the {\tt CircuitPiece} 
region relationship ensures that {\tt rn0 $\leq$ rn}.
